Step 13: Marginal Effects Analysis with Multiple Imputation

### 2025-09-08 reanalysis-13-marginal_effects.qmd
ame

```{r define-subgroups}
# Define the subgroups of interest (matching analysis.Rmd)
subgroups <- list(
  overall = list(
    name = "Overall (Average Marginal Effect)",
    subset_condition = NULL,  # No subsetting for overall
    description = "Average marginal effect across entire sample"
  ),

  pdr_any_tx = list(
    name = "PDR on any treatment",
    subset_condition = expression(PDR == "Present" & 
                                 (PRP_flag == 1 | anti_VEGF == 1 | 
                                  focal_laser_flag == 1 | other_inject == 1)),
    description = "Patients with PDR receiving any treatment"
  ),

  pdr_anti_vegf_only = list(
    name = "PDR on anti-VEGF only",
    subset_condition = expression(PDR == "Present" & 
                                 PRP_flag == 0 & anti_VEGF == 1 & 
                                 focal_laser_flag == 0 & other_inject == 0),
    description = "Patients with PDR receiving anti-VEGF only"
  ),

  pdr_prp_only = list(
    name = "PDR on PRP only",
    subset_condition = expression(PDR == "Present" & 
                                 PRP_flag == 1 & anti_VEGF == 0 & 
                                 focal_laser_flag == 0 & other_inject == 0),
    description = "Patients with PDR receiving PRP only"
  )
)

@@

>>>>

```{r define-subgroups}
# Define all subgroups from twang-analysis.Rmd
subgroups <- list(
  # Overall population
  overall = list(
    name = "Overall Population",
    subset_condition = NULL,  # No subsetting for overall
    description = "Average marginal effect across entire sample"
  ),

  # PDR subgroups - using dr_severity == "PDR"
  pdr_any_tx = list(
    name = "PDR, any treatment",
    subset_condition = expression(dr_severity == "PDR" & 
                                 (PRP_flag == 1 | anti_VEGF == 1 | 
                                  focal_laser_flag == 1 | other_inject == 1)),
    description = "Patients with PDR receiving any treatment"
  ),

  pdr_anti_vegf_only = list(
    name = "PDR, on anti-VEGF only",
    subset_condition = expression(dr_severity == "PDR" & 
                                 PRP_flag == 0 & anti_VEGF == 1 & 
                                 focal_laser_flag == 0 & other_inject == 0),
    description = "Patients with PDR receiving anti-VEGF only"
  ),

  pdr_prp_only = list(
    name = "PDR, on PRP only",
    subset_condition = expression(dr_severity == "PDR" & 
                                 PRP_flag == 1 & anti_VEGF == 0 & 
                                 focal_laser_flag == 0 & other_inject == 0),
    description = "Patients with PDR receiving PRP only"
  ),

  # No DR or NPDR combined groups - using dr_severity levels
  npdr_nodr_no_tx = list(
    name = "No DR or NPDR, no treatment",
    subset_condition = expression(dr_severity %in% c("No_DR", "NPDR") & 
                                 (PRP_flag == 0 & anti_VEGF == 0 & 
                                  focal_laser_flag == 0 & other_inject == 0)),
    description = "Patients with No DR or NPDR not on treatment"
  ),

  npdr_nodr_any_tx = list(
    name = "No DR or NPDR, any treatment",
    subset_condition = expression(dr_severity %in% c("No_DR", "NPDR") & 
                                 (PRP_flag == 1 | anti_VEGF == 1 | 
                                  focal_laser_flag == 1 | other_inject == 1)),
    description = "Patients with No DR or NPDR on any treatment"
  ),

  # NPDR overall groups - using dr_severity == "NPDR"
  npdr_no_tx = list(
    name = "NPDR, no treatment",
    subset_condition = expression(dr_severity == "NPDR" & 
                                 (PRP_flag == 0 & anti_VEGF == 0 & 
                                  focal_laser_flag == 0 & other_inject == 0)),
    description = "Patients with NPDR not on treatment"
  ),

  npdr_any_tx = list(
    name = "NPDR, any treatment",
    subset_condition = expression(dr_severity == "NPDR" & 
                                 (PRP_flag == 1 | anti_VEGF == 1 | 
                                  focal_laser_flag == 1 | other_inject == 1)),
    description = "Patients with NPDR on any treatment"
  )

  # Note: NPDR severity-specific subgroups (Mild/Moderate/Severe) removed
  # because dr_severity was imputed at coarse level (No_DR/NPDR/PDR only)
)

@@

## Forest Plots

```{r forest-plots}
#| fig.width: 10
#| fig.height: 8

# Prepare data for forest plot
forest_data <- do.call(rbind, lapply(names(pooled_mfx), function(sg_name) {
  res <- pooled_mfx[[sg_name]]
  if (res$success) {
    data.frame(
      label = res$name,
      AME = res$ame,
      SE = res$se,
      Lower = res$ci_lower,
      Upper = res$ci_upper,
      p = res$p_value,
      n = res$avg_n,
      group_type = ifelse(sg_name == "overall", "overall", "subgroup"),
      stringsAsFactors = FALSE
    )
  }
}))

# Add sample sizes to labels
forest_data$label_with_n <- paste0(forest_data$label, " (n=", round(forest_data$n), ")")

# Define order for plots (reversed for coord_flip)
subgroup_order <- rev(c(
  "No DR or NPDR, no treatment", "No DR or NPDR, any treatment", 
  "NPDR, no treatment", "NPDR, any treatment", 
  "Mild NPDR, no treatment", "Mild NPDR, any treatment",
  "Moderate NPDR, no treatment", "Moderate NPDR, any treatment",
  "Severe NPDR, no treatment", "Severe NPDR, any treatment",
  "PDR, any treatment", "PDR, on anti-VEGF only", "PDR, on PRP only"
))

# Version 1: Subgroups only
data_subgroups <- forest_data[forest_data$group_type == "subgroup", ]
data_subgroups$Labels <- factor(data_subgroups$label, levels = subgroup_order)

plot_subgroups <- ggplot(data_subgroups, aes(x = AME, y = Labels)) +
  geom_vline(xintercept = 0, color = "#be123c", linetype = "solid", size = 0.5) +
  geom_point(size = 3, color = "#2563eb") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2, color = "#2563eb") +
  geom_text(
    aes(x = Lower, label = sprintf("%.3f", Lower)), 
    hjust = 1.2, size = 3, color = "black"
  ) +
  geom_text(
    aes(x = Upper, label = sprintf("%.3f", Upper)), 
    hjust = -0.2, size = 3, color = "black"
  ) +
  geom_text(
    aes(label = sprintf("%.3f", AME)), 
    hjust = -0.2, vjust = -1, size = 3, color = "black"
  ) +
  theme_minimal() +
  scale_x_continuous(
    limits = c(-0.01, max(forest_data$Upper) * 1.1),
    breaks = scales::pretty_breaks(n = 6)
  ) +
  labs(
    title = "Average Marginal Effects of Lapse Across Selected Groups",
    subtitle = "Pooled across multiple imputations using Rubin's rules",
    x = "Average Marginal Effect (AME)",
    y = "Treatment Group",
    caption = paste0("Based on ", unique(forest_data$n_imputations)[1], " imputations")
  ) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10), 
    axis.title.y = element_text(size = 12, face = "bold", margin = ggplot2::margin(r = 20)),
    axis.title.x = element_text(size = 12, face = "bold", margin = ggplot2::margin(t = 20)),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = ggplot2::margin(b = 20)),
    plot.subtitle = element_text(size = 11, hjust = 0.5, margin = ggplot2::margin(b = 10)),
    plot.margin = ggplot2::margin(10, 20, 30, 20),
    panel.grid.major.y = element_line(color = "gray", linetype = "dotted"),
    plot.caption = element_text(size = 10, face = "italic")
  )

# Version 2: With overall population included
all_order <- c("Overall Population", subgroup_order)
forest_data$Labels <- factor(forest_data$label, levels = rev(all_order))
forest_data$point_color <- ifelse(forest_data$group_type == "overall", "#dc2626", "#2563eb")
forest_data$point_size <- ifelse(forest_data$group_type == "overall", 4, 3)

plot_with_overall <- ggplot(forest_data, aes(x = AME, y = Labels)) +
  geom_vline(xintercept = 0, color = "#be123c", linetype = "solid", size = 0.5) +
  geom_hline(yintercept = length(subgroup_order) + 0.5, 
             color = "gray50", linetype = "dashed", alpha = 0.5) +
  geom_point(aes(color = point_color, size = point_size)) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper, color = point_color), height = 0.2) +
  geom_text(
    aes(x = Lower, label = sprintf("%.3f", Lower)), 
    hjust = 1.2, size = 3, color = "black"
  ) +
  geom_text(
    aes(x = Upper, label = sprintf("%.3f", Upper)), 
    hjust = -0.2, size = 3, color = "black"
  ) +
  geom_text(
    aes(label = sprintf("%.3f", AME)), 
    hjust = -0.2, vjust = -1, size = 3, color = "black"
  ) +
  scale_color_identity() +
  scale_size_identity() +
  theme_minimal() +
  scale_x_continuous(
    limits = c(-0.01, max(forest_data$Upper) * 1.1),
    breaks = scales::pretty_breaks(n = 6)
  ) +
  labs(
    title = "Average Marginal Effects of Lapse Across All Groups",
    subtitle = "Pooled across multiple imputations using Rubin's rules",
    x = "Average Marginal Effect (AME)",
    y = "Treatment Group",
    caption = paste0("Based on ", unique(forest_data$n_imputations)[1], " imputations\n",
                    "Overall population shown in red, subgroups in blue")
  ) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10), 
    axis.title.y = element_text(size = 12, face = "bold", margin = ggplot2::margin(r = 20)),
    axis.title.x = element_text(size = 12, face = "bold", margin = ggplot2::margin(t = 20)),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = ggplot2::margin(b = 20)),
    plot.subtitle = element_text(size = 11, hjust = 0.5, margin = ggplot2::margin(b = 10)),
    plot.margin = ggplot2::margin(10, 20, 30, 20),
    panel.grid.major.y = element_line(color = "gray", linetype = "dotted"),
    plot.caption = element_text(size = 10, face = "italic")
  )

# Display plots
print(plot_subgroups)
print(plot_with_overall)

# Save plots
ggsave(file.path(marginal_effects_dir, "forest_plot_subgroups_only.png"), 
       plot_subgroups, width = 10, height = 8, dpi = 300, bg = "white")
ggsave(file.path(marginal_effects_dir, "forest_plot_with_overall.png"), 
       plot_with_overall, width = 10, height = 8.5, dpi = 300, bg = "white")

log_message("Forest plots saved")
```
